---
title: "Introduction to AbNames"
author: "Helen Lindsay"
package: AbNames
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{AbNames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Overview

AbNames performs two tasks:

* Extracting potential gene names from antibody names, and

* Matching human antibody names to gene names and IDs.

Antibody names are not reported in a consistent format.  Antibodies are often
named according to the antigen they target, which may not be the same as the
name of the protein (complex) the antigen is part of.  Antibodies may target 
multi-subunit protein complexes, and this can be reflected in the name, e.g. an
antibody against the T-cell receptor alpha and beta subunits might be named
TCRab. Antibody names may also include the name of the clone the antibody is
derived from or the names of fluorophores or DNA-oligos the antibody is
conjugated with.  For these reasons, it can be difficult to exactly match
antibody names with gene or protein names.  As cell surface antigens often have
very similar names, searching for partial matches to names in free-text gene descriptions is challenging and error-prone.

# Antibody naming conventions

## T-cell receptors

## Immunoglobins

## Major histocompatibility complex genes 

## Isotype controls


# Data

AbNames contains several curated gene name data sets for matching to antibody
names:

## HGNC gene names

AbNames contains a reformatted version of the protein-coding gene information
downloaded from the Human Gene Names Consortium. This includes previous
(obsolete) gene names and aliases, which in our experience have been useful for 
matching to antibody names

## CITE-seq antibodies

This is a table matching antibody names to gene and protein IDs from >20 data
sets with publicly available CITE-seq data.  The AbNames package collects the
functions that were used to create this table.  This table includes manually
curated matches between antibody names and gene IDs for cases where we were
unable to find an exact match to the antibody name provided.

## BioLegend antibodies

TO DO: CHECK WITH BIOLEGEND IF GENE NAME DATA CAN BE INCLUDED IN THIS PACKAGE
BioLegend is a major supplier of antibodies, and provides several antibody
panels for CITE-seq analyses.  


# Creating a query table

```{r setup}
library(AbNames)
```

## Default query table

We will demonstrate how to create a default query table using the raw CITE-seq
data set.  This collects information from the reagents tables provided as supplementary material in the studies used.

First, we load the unnannotated data table:

```{r}
citeseq_fname <- system.file("extdata", "citeseq.csv", package = "AbNames")
citeseq <- read.csv(citeseq_fname, sep = ",")

```


## Custom query table

AbNames provides a few template functions that can be used to construct a pipeline for creating a query table.  


### Special case: T-cell receptors

We found that querying the HGNC gene description was the easiest way to match
the names of antibodies against the T-cell receptor to their gene IDs.  (An
alternative method is to search for the gene symbol.)

We will start with a data.frame of antibodies against the T-cell receptor from
the CITE-seq data set. We have already done some fomatting of these names, e.g. 
for "TCR Va24-Ja18 (iNKT cell)" we have removed the section in brackets. We wish
to create a query data.frame where each subunit of the TCR complex appears on a
separate line. 

```{r}
tcr <- data.frame(Antigen =
                      c("TCR alpha/beta", "TCRab", "TCR gamma/delta", "TCRgd",
                        "TCR g/d", "TCR Vgamma9", "TCR Vg9", "TCR Vd2",
                        "TCR Vdelta2", "TCR Vα24-Jα18", "TCRVa24.Ja18", 
                        "TCR Valpha24-Jalpha18", "TCR Vα7.2", "TCR Va7.2",
                        "TCRa7.2", "TCRVa7.2", "TCR Vbeta13.1", "TCR γ/δ", 
                        "TCR Vβ13.1", "TCR Vγ9", "TCR Vδ2", "TCR α/β",
                        "TCRb", "TCRg"))


# First, we convert the Greek symbols to letters.  
# Note that as we are using replaceGreekSyms in a dplyr pipeline, we don't 
# put quotes around the column name, i.e. Antigen not "Antigen".

tcr <- tcr %>%
    dplyr::mutate(query = 
                    AbNames::replaceGreekSyms(Antigen, replace = "sym2letter"))
 
# Print out a few rows to see the result
tcr %>% 
    dplyr::filter(Antigen %in% c("TCR Vβ13.1", "TCR Vδ2", "TCR α/β"))

```

Now we use the function `formatTCR` to format the antibody names for querying
the gene description field of the HGNC data set.

```{r}
tcr_f <- AbNames::formatTCR(tcr, tcr = "query")

# Print out the first few rows
tcr_f %>%
    head() 

```


# Querying a dataset


# Querying the protein ontology


# Filling NAs from a reference

When filling in information from a reference data set, it can be useful to look
at entries where for example annotations are inconsistent.  `r showGroups` is
a function that allows users to interactively print group(s) from a grouped
data.frame.

