---
title: "Introduction to AbNames"
author: "Helen Lindsay"
package: AbNames
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{AbNames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Overview

AbNames performs two tasks:

* Extracting potential gene names from antibody names, and

* Matching human antibody names to gene names and IDs.

Antibody names are not reported in a consistent format.  Antibodies are often
named according to the antigen they target, which may not be the same as the
name of the protein (complex) the antigen is part of.  Antibodies may target 
multi-subunit protein complexes, and this can be reflected in the name, e.g. an
antibody against the T-cell receptor alpha and beta subunits might be named
TCRab. Antibody names may also include the name of the clone the antibody is
derived from or the names of fluorophores or DNA-oligos the antibody is
conjugated with.  For these reasons, it can be difficult to exactly match
antibody names with gene or protein names.  As cell surface antigens often have
very similar names, searching for partial matches to names in free-text gene descriptions is challenging and error-prone.

# Antibody naming conventions

## T-cell receptors

## Immunoglobins

## Major histocompatibility complex genes 

## Isotype controls


# Data

AbNames contains several curated gene name data sets for matching to antibody
names:

## HGNC gene names

AbNames contains a reformatted version of the protein-coding gene information
downloaded from the Human Gene Names Consortium. This includes previous
(obsolete) gene names and aliases, which in our experience have been useful for 
matching to antibody names

## CITE-seq antibodies

This is a table matching antibody names to gene and protein IDs from >20 data
sets with publicly available CITE-seq data.  The AbNames package collects the
functions that were used to create this table.  This table includes manually
curated matches between antibody names and gene IDs for cases where we were
unable to find an exact match to the antibody name provided.

## BioLegend antibodies

TO DO: CHECK WITH BIOLEGEND IF GENE NAME DATA CAN BE INCLUDED IN THIS PACKAGE
BioLegend is a major supplier of antibodies, and provides several antibody
panels for CITE-seq analyses.  


# Creating a query table

```{r setup}
library(AbNames)
```

## Default query table


## Custom query table

### Special case: T-cell receptors

We found that querying the HGNC gene description was the easiest way to match
the names of antibodies against the T-cell receptor to their gene IDs.  (An
alternative method is to search for the gene symbol.)

We will start with a data.frame of antibodies against the T-cell receptor from
the CITE-seq data set. 

```{r}
tcr <- data.frame(Antigen =
                      c("TCR alpha/beta", "TCR alpha/beta (TRAC/TRBC)",
                        "TCRab", "TCR gamma/delta", "TCRgd", "TCR g/d",
                        "TCR Vgamma9", "TCR Vg9", "TCR Vd2", "TCR Vdelta2",
                        "TCR Vα24-Jα18", "TCR Va24-Ja18 (iNKT cell)",
                        "TCRVa24.Ja18", "TCR Va24-Ja18",
                        "TCR Valpha24-Jalpha18 (iNKT cell)",
                        "TCR Vα7.2", "TCR Va7.2", "TCRa7.2", "TCRVa7.2",
                        "TCR Vbeta13.1", "TCR γ/δ", "TCR Vα24-Jα18 (iNKT cell)",
                        "TCR Vβ13.1", "TCR Vγ9", "TCR Vδ2", "TCR α/β",
                        "TCRb", "TCRg"))

```

We wish to create a data.frame where each subunit of the TCR complex appears on
a separate line.


# Querying a dataset


# Querying the protein ontology
