---
title: "Introduction to AbNames"
author: "Helen Lindsay"
package: AbNames
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{AbNames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Overview

AbNames performs two tasks:

* Extracting potential gene names from antibody names, and

* Matching human antibody names to gene names, IDs and protein complex IDs.

Antibody names are not reported in a consistent format in published data.  Antibodies are often named according to the antigen they target, which may not
be the same as the name of the protein (complex) the antigen is part of. Antibodies may target multi-subunit protein complexes, and this can be reflected
in the name, e.g. an antibody against the T-cell receptor alpha and beta subunits might be named TCRab. Antibody names may also include the name of the clone the antibody is derived from or the names of fluorophores or DNA-oligos the antibody
is conjugated with.  For these reasons, it can be difficult to exactly match
antibody names with gene or protein names.  As cell surface antigens often have
very similar names, searching for partial matches to names in free-text gene descriptions is challenging and error-prone.

# Goals

 Standardise names of data sets for combined analysis

## Identify all members of a protein complex bound by an antibody

# Data

AbNames contains several curated gene name data sets for matching to antibody
names.

## HGNC gene names

The `hgnc` data set is a reformatted version of the protein-coding gene
information downloaded from the Human Gene Names Consortium. This includes
previous (obsolete) gene names and aliases, which in our experience have been
useful for matching to antibody names.  The `hgnc_long` data set contains the
same seq of gene identifiers as the `hgnc` data set, in long format (one alias
per row).

Load using:

```{r, message = FALSE}
library(AbNames)
library(dplyr)

data("hgnc", package = "AbNames") # hgnc_long is loaded similarly

# Show the first entries of hgnc, where each row is the start of one column
dplyr::glimpse(hgnc) 

# (Note: it isn't necessary to use dplyr:: to call "glimpse" as dplyr is loaded
# with the library call above. This syntax is used to make it clear which
# packages functions belong to)
```

## BioLegend antibodies

BioLegend is a major supplier of antibodies, and provides several antibody
panels for CITE-seq analyses.  The data set "totalseq_cocktails" is a
re-formatted version of the TotalSeq cocktails data sheets available from the
BioLegend website, including BioLegend antibody names and Ensembl gene IDs.  The isotypes of the antibodies are not included in the TotalSeq data sheets.

Load using:
```{r}
data("totalseq_cocktails", package = "AbNames")
dplyr::glimpse(totalseq_cocktails)
```

Note that the "Antigen" column here refers to antibody names with prefixes such
as "anti-human" removed.

## CITE-seq antibodies

This is a table matching antibody names to gene and protein IDs from >20 data
sets with publicly available CITE-seq data.  The data sets are data that we have worked with and we would be happy to add other data sets if provided. The
AbNames package collects the functions that were used to create this table. This table (NOT IN THERE YET) includes manually curated matches between antibody names and gene IDs for cases where we were unable to find an exact match to the antibody name provided.  Isotype controls were manually identified based on either the 
name, e.g. "IgG1 Isotype Ctrl" or the reactivity, e.g. "Mouse IgG2b".

Load using
```{r, message = FALSE}
# As the citeseq data set contains raw data, it is loaded differently
# than the other data sets 

citeseq_fname <- system.file("extdata", "citeseq.csv", package = "AbNames")
citeseq <- read.csv(citeseq_fname) %>% unique()
dplyr::glimpse(citeseq)

# TO DO: AS THIS IS ONLY A SUBSET OF THE CITESEQ INFO, THERE ARE APPARENT
# DUPLICATES, DECIDE WHAT TO DO WITH THESE
```

Here the column "Antigen" refers to the name given by the authors of the study
in the table of resources used.  Some minimal formatting has been done, for
example fixing Greek characters that were accidentally transformed upon data import.

# Creating a query table

To illustrate the problem with matching antibody names to gene names, let's look
at an example from the citeseq data set.

```{r}
# The regular expression inside "grepl" searches for PD.L1, PD-L1, PDL1 or CD274

cd274 <- citeseq %>%
    dplyr::filter(grepl("PD[\\.-]?L1|CD274", Antigen)) %>%
    dplyr::pull(Antigen) %>%
    unique()

cd274
```

All of the antigen names above refer to the same cell surface protein. 

## Default query table

We will demonstrate how to create a default query table using the raw CITE-seq
data set.  This collects information from the reagents tables provided as supplementary material in the studies used.


```{r}
# Add an ID column to the citeseq data set to allow the results table to be merged
citeseq <- AbNames::addID(citeseq)

# Remove control columns, as we are only searching human proteins
controls <- dplyr::filter(citeseq, Control)
citeseq <- dplyr::filter(citeseq, ! Control)

# Select just the columns that are needed for querying:
citeseq_q <- citeseq %>% dplyr::select(ID, Antigen)

# Apply the default transformation sequence and make a query table in long format
query_df <- AbNames::makeQueryTable(citeseq_q, ab = "Antigen")

# Print one example antigen from the query table:
query_df %>%
    dplyr::filter(ID == "TCR alpha/beta__Hao_2021")

```

The example above shows that each antibody name has been reformatted in several
ways.  When querying a data set, we search for an exact match to any of these
strings.

## Custom query table

AbNames provides a few template functions that can be used to construct a
pipeline for creating a query table. The default query table uses the function
`defaultQuery` to create a list of partial functions which are then applied
recursively to the data.frame using `magrittr::freduce`.  To use this strategy,
all functions must accept the data.frame as the only required argument.  Other
arguments should be filled in when creating the partial functions.

The default function list can used as a starting point to add extra formatting
functions, remove or modify certain steps.  The individual formatting functions
are introduced below (SO FAR ONLY THE T-CELL RECEPTORS)

```{r}
# Get the default sequence of formatting functions 
default_funs <- AbNames::defaultQuery()

# Print the first two formatting functions as an example
default_funs[1:2]
```


### T-cell receptors

We found that querying the HGNC gene description was the easiest way to match
the names of antibodies against the T-cell receptor to their gene IDs.  (An
alternative method is to search for the gene symbol.)

We will start with a data.frame of antibodies against the T-cell receptor from
the CITE-seq data set. We have already done some fomatting of these names, e.g. 
for "TCR Va24-Ja18 (iNKT cell)" we have removed the section in brackets. We wish
to create a query data.frame where each subunit of the TCR complex appears on a
separate line. 

```{r}
tcr <- data.frame(Antigen =
                      c("TCR alpha/beta", "TCRab", "TCR gamma/delta", "TCRgd",
                        "TCR g/d", "TCR Vgamma9", "TCR Vg9", "TCR Vd2",
                        "TCR Vdelta2", "TCR Vα24-Jα18", "TCRVa24.Ja18", 
                        "TCR Valpha24-Jalpha18", "TCR Vα7.2", "TCR Va7.2",
                        "TCRa7.2", "TCRVa7.2", "TCR Vbeta13.1", "TCR γ/δ", 
                        "TCR Vβ13.1", "TCR Vγ9", "TCR Vδ2", "TCR α/β",
                        "TCRb", "TCRg"))


# First, we convert the Greek symbols to letters.  
# Note that as we are using replaceGreekSyms in a dplyr pipeline, we don't 
# put quotes around the column name, i.e. Antigen not "Antigen".

tcr <- tcr %>%
    dplyr::mutate(query = 
                    AbNames::replaceGreekSyms(Antigen, replace = "sym2letter"))
 
# Print out a few rows to see the result
tcr %>% 
    dplyr::filter(Antigen %in% c("TCR Vβ13.1", "TCR Vδ2", "TCR α/β"))

```

Now we use the function `formatTCR` to format the antibody names for querying
the gene description field of the HGNC data set.

```{r}
tcr_f <- AbNames::formatTCR(tcr, tcr = "query")

# Print out the first few rows
tcr_f %>%
    head() 

```


# Querying a dataset

Our pipeline is to query the HGNC and then use the other datasets and
the antibody vendor information to find matches for the unmatched antibodies.

By default, we require that matches must be found for all subunits of a
multi-subunit protein to avoid incorrect matches.

Here we will query the HGNC data set for the CITE-seq antibodies, using the
query table created above.  

```{r, echo = FALSE}
set.seed(549867)
```

```{r}
hgnc_results <- searchHGNC(query_df)

# Print 10 random results:
hgnc_results %>%
    dplyr::select(ID, name, value, symbol_type) %>%
    dplyr::ungroup() %>%
    dplyr::sample_n(10)
```

The results table above contains (just) the matches between the query table and the HGNC tables.  In the example above, we see in the "name" column the formatting function that generated the string in the "value" column that was matched, and in the "symbol_type" column which column of the HGNC data was matched.  If there are multiple matches for a given ID, the official symbol ("HGNC_SYMBOL") is preferred over aliases or previous symbols.

We may wish to review the matches before merging the results into the original table.
For example, we can check for matches to different genes where the name was not guessed to be a multi-subunit protein.

```{r}
hgnc_results <- hgnc_results %>%
    dplyr::group_by(ID) %>%
    dplyr::mutate(n_ids = dplyr::n_distinct(HGNC_ID)) # Count distinct IDs
    
# Select antigens where there are matches to multiple genes but not
# because the antibody is against a multi-gene protein 
multi_gene <- hgnc_results %>%
    dplyr::filter(n_ids > 1, ! all(name %in% c("TCR_long", "subunit"))) %>%
    dplyr::select(ID, name, value, HGNC_ID)

# Look at the first group in multi_gene.
# Set interactive = FALSE for interactive exploration
showGroups(multi_gene, 1, interactive = FALSE)

# This is an example where the antibody is against a heterodimeric protein.
# We can confirm this by looking up the vendor catalogue number:
citeseq %>%
    dplyr::filter(ID == "CD11a.CD18-Wu_2021_b") %>%
    dplyr::select(Antigen, Cat_Number, Vendor)

```

However, there are also cases where the results table contains false matches, for example:

```{r}
hgnc_results %>%
    dplyr::filter(ID == "CD270 (HVEM, TR2)-Hao_2021") %>%
    data.frame()
```

In the above table, we see from the "value" column that the gene "ENSG00000157873" has alias symbols "CD270", "HVEM" and "TR2".  The next two
rows show genes that have alias symbols for only "TR2".  Here we could guess
that the gene with the most matches is the correct one. 

NOTE: I haven't written a convenience function for merging back into the original table yet.
TODO:

* Add official HGNC symbols back into to the HGNC data set
* QUERY FOR CD11/CD18 in protein ontology
* Remove redundant results because of Antigen/greek_letter e.g. "KLRG1 (MAFA)-Qian_2020" 

Other examples to discuss?

*"KIR2DL5" matches two genes HGNC:16345 HGNC:16346

*"CD279 (PD-1)-Mimitou_2021_cite_and_dogma_seq"

*"DR3 (TRAMP)-Liu_2021"

For now I will remove all genes with multiple matches before merging the results into the citeseq data.

```{r}
nrow(hgnc_results)

# Remove matches to several genes, select just columns of interest
hgnc_results <- hgnc_results %>%
    dplyr::filter(n_ids == 1 | all(name %in% c("TCR_long", "subunit"))) %>%
    dplyr::select(ID, HGNC_ID, HGNC_SYMBOL, ENSEMBL_ID, UNIPROT_IDS) %>%
    unique() %>% # Collapse results with same ID from different queries
    dplyr::summarise(dplyr::across(c(HGNC_ID, HGNC_SYMBOL,
                                     ENSEMBL_ID, UNIPROT_IDS),
                                   toString)) # Collapse multi-subunit entries

nrow(hgnc_results)

citeseq <- citeseq %>%
    dplyr::left_join(hgnc_results, by = "ID") %>%
    dplyr::relocate(ID, Antigen, Cat_Number, HGNC_ID) %>%
    unique()

head(citeseq)
```

# Filling missing information

## Filling using the TotalSeq cocktails

We can annotate some of the antibodies that could not be annotated using the
HGNC by using the totalseq_cocktails table:

```{r}
count_missing <- function(df){
    dplyr::filter(df, is.na(HGNC_ID)) %>% nrow()
}

data(totalseq_cocktails)
ts <- totalseq_cocktails %>%
    dplyr::select(any_of(colnames(citeseq)))

missing <- count_missing(citeseq)

# Fill in IDs where Antigen, Oligo, Clone and TotalSeq category match
citeseq <- citeseq %>%
    dplyr::rows_patch(ts,
                      by = c("Antigen", "Oligo_ID", "Clone", "TotalSeq_Cat"),
                      unmatched = "ignore")

missing_after_ts <- count_missing(citeseq)

# Missing before filling:
missing

# Still missing after filling with TotalSeq:
missing_after_ts
```


## Inspecting groups before filling

When filling in information from a reference data set, it can be useful to look
at entries where for example annotations are inconsistent.  `showGroups` is
a function that allows users to interactively print group(s) from a grouped
data.frame.  We give a (non-interactive) example below.

## Filling NAs using a reference data set

The function `fillByGroup` is used to fill in NAs in a grouped data.frame.  It differs from `tidyr::fill` in its treatment of inconsistent values.  Whereas `tidyr::fill` will fill using the first value, `fillByGroup` offers the option
to fill in the most frequent value.  This can be useful when filling antibody
IDs given the antibody name and clone or catalogue number.

To fill using a reference data set, the strategy used is either to add a
temporary ID column, join the two data.frames, fill and separate again using the
ID column, or use dplyr::rows_patch.  We will demonstrate the latter approach below. 

```{r}
# Before filling, check how many antibodies were matched.
original_nmatched <- count_missing(citeseq)

# Select some data to demonstrate filling:
# Get entries sharing the same catalogue number,
# where not every entry has a match
fill_demo <- citeseq %>%
    dplyr::group_by(Cat_Number) %>%
    dplyr::arrange(Cat_Number) %>%
    dplyr::filter(!is.na(Cat_Number), 
                  any(is.na(HGNC_ID)),
                  ! all(is.na(HGNC_ID)))

AbNames::showGroups(fill_demo, interactive = FALSE)
```

In the above example, we can see that for antibodies with catalogue number "300475", a match was found if the antibody was named "CD3E" but not if it was named "CD3".  We can fill in the missing information using `fillByGroup`:

```{r}
fill_demo <- AbNames::fillByGroup(fill_demo, "Cat_Number",
                                  fill = c("HGNC_ID", "TotalSeq_Cat", "Vendor",
                                           "ENSEMBL_ID", "UNIPROT_IDS")) %>%
    dplyr::group_by(Cat_Number) # Re-group as fillByGroup ungroups

# Print out the first group again
AbNames::showGroups(fill_demo, interactive = FALSE)
```

Now we will fill the gene information in the citeseq data similarly, and check
how many antibodies have not been matched.

```{r}
# We fill by grouping the Catalogue number:
citeseq <- citeseq %>%
     AbNames::fillByGroup("Cat_Number", multiple = "mode",
                          fill = c("HGNC_ID", "TotalSeq_Cat", "Vendor",
                                           "ENSEMBL_ID", "UNIPROT_IDS"))

nmatched_after_fill <- count_missing(citeseq)

print("Before filling:")
original_nmatched
print("After filling:")
nmatched_after_fill


```





# Querying the protein ontology

Let's see which antibodies were not matched in the HGNC.

```{r}
citeseq %>%
    dplyr::filter(is.na(HGNC_ID)) %>%
    pull(Antigen) %>%
    sort() %>%
    unique()

missing <- citeseq %>%
    dplyr::filter(is.na(HGNC_ID)) %>%
    dplyr::select(ID, Antigen)
```


## TO DO

*An example of (interactively) deciding what to do if groups are inconsistent.
*Standardise names in a singlecellexperiment
*Check new annotations against previous
*Isotype controls
*Add checks on filling if the values are not unique
*Add control column to isotype controls to prevent false match to mouse IgG2a
*Add ignore option to split merge?
* CD25 (good example, multiple genes), CD270, CD279 matched in HGNC?
# CD279 PD-1 - example where only some are matched
# Fix Hao Clone CD25
# Search for similar genes in master table
# Generate camel case regexp?
# Split HGNC alias_name, prev_name - keep only if contains "antigen" or "CD"?
# Unintuitive to use quoted for splitUnnest
# Uniprot ids - possible to split to create complex uniprot and modification
# uniprot?
# Redocument data sets
# HAO CD45 should have catalogue numbers!
# union join back gene ID / get info by antigen


# Session Info

```{r}
sessionInfo()
```
