---
title: "Standardize a SingleCellExperiment"
author: "Helen Lindsay"
package: AbNames
date: '2022-07-11'
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{AbNames_SCE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This vignette shows how to standardize the names of the antibody-derived tags in
a CITE-seq experiment in SingleCellExperiment format.

To demonstrate, we will use the "Kotliarov" dataset available in the package
"scRNAseq".  


```{r, message = FALSE}

# May not need much renaming....
library("scRNAseq")
library("AbNames")

kotliarov <- KotliarovPBMCData(mode = "adt")
head(rownames(kotliarov))
```

## Put into a data frame to allow matching to the CITEseq data set

You can include information such as catalogue number and antibody clone
to assist matching if it is available.  When matching to the CITEseq data
set, the names of the columns to use for matching must equal those in the
CITEseq data set.  For example, below we match using only the antigen name,
using the column "Antigen".  By default, columns "Clone", "Antigen", and "Cat_Number" (catalogue number) are used for matching if available. 

# HGNC_ID


```{r}
df <- data.frame(Original = rownames(kotliarov),
                 Antigen = gsub("_PROT", "", rownames(kotliarov)))


```
## Match to the CITEseq data set and review suggested consensus names

It is a good idea to review matches as errors in e.g. catalogue number will lead to incorrect matches.

The function `matchToCiteseq` will report which columns are used for matching.
Note that unless specified otherwise, the entries of the CITEseq data set are matched using the default columns even if these do not exist in the new data set.


```{r}
df <- matchToCiteseq(df)
print(head(df))
```


## Rename single cell experiment
```{r}

# To rename the singleCellExperiment, we need a named vector of the new names,
# with names being the original names of the singleCellExperiment.
new_nms <- structure(df[[2]], names = rownames(kotliarov))
print(head(new_nms))

kotliarov <- renameADT(kotliarov, new_nms)

head(rownames(kotliarov))



```


```{r}
sessionInfo()
```

```{r}
# alternative data sets
#library(SingleCellMultiModal)
#cord_blood <- CITEseq(DataType="cord_blood", modes="scADT_Counts",
#                      dry.run=FALSE, DataClass = "SingleCellExperiment")
#cord_blood

#library(CiteFuse)
#data("CITEseq_example", package = "CiteFuse")
#sce_citeseq <- preprocessing(CITEseq_example)
#rownames(altExp(sce_citeseq, "ADT"))
```{r}
